\textbf{Efficient Implementation Strategies for Block Ciphers on ARMv8}\\ {\footnotesize Bachelorarbeit}

### 2023-05-17 Paper/chapters/implementation.tex
camellia galois affine isomorphism
add macros to paper

@@

\subsection{Bitslicing}

We will examine the round function in closer detail and compare the source code
with the generated assembly.
\pagebreak

\begin{lstlisting}[language=c, caption={Round function src}]
for (int round = 0; round < ROUNDS_GIFT_64; round++) {
    gift_64_vec_sliced_subcells(s);
    gift_64_vec_sliced_permute(s);

@@

Without needing to dive too deep into details, it is obvious to see the
compiler has inlined the two function calls to \texttt{subcells} and
\texttt{permute} with lines 1 to 23 originating from the \texttt{subcells} and
lines 24-46 from the \texttt{permute} and round key addition functions. It has
chosen not to unroll the loop, but has moved the loop counter increment as well
as the condition check in between the \texttt{subcells} instructions to line
\ref{lst:cntrincr} and \ref{lst:cntrchk}. In addition, the loop counter serves
a second purpose as an offset register and is therefore incremented by 0x80=128
instead of just 1.

>>>>

Lots of operations require extraction of a certain number of consecutive bits,
usually referred to as a bitfield. Indices for table lookups are generally
attained by right-shifting a larger value stored in a register, then applying
an AND operation to get the lowest $n$ bits and finally writing the result into
the beginning of another register. Due to its usefulness, this operation is
implemented as \texttt{UBFX} for an unsigned bitfield extract and can be used
to implement S-box extraction for subperm lookups which would otherwise take
two or three instructions.

@@

\subsection{Bitslicing}
\subsubsection{Shifts for swapmove}

Implementing shift functions for 128-bit NEON registers by extracting the
overflow and adding it back in later can be realized using 5 instructions. It
would be most useful for the function to take a shift amount parameter, but
most NEON intrinsics encode their parameters into the final machine code such
that parameters need to be compile-time constants. We will therefore implement
\texttt{shl}, \texttt{shr} and \texttt{swapmove} as C macros utilizing the
\texttt{vextq\_u64} intrinsic to swap the two 64-bit vector elements:

\begin{lstlisting}{language=c, caption={Shift and swapmove}}

/*
uint8x16_t shl(const uint8x16_t v, const int n) */
#define shl(_a, v, n)                                                       \
{                                                                           \
        uint64x2_t _overflow = vshrq_n_u64(v, 64 - n);                      \
        _overflow = vextq_u64(vdupq_n_u64(0x0), _overflow, 1);              \
        _a = vorrq_u8(vshlq_n_u64(v, n), _overflow);                        \
}

/*
uint8x16_t shr(const uint8x16_t v, const int n) */
#define shr(_a, v, n)                                                       \
{                                                                           \
        uint64x2_t _overflow = vshlq_n_u64(v, 64 - n);                      \
        _overflow = vextq_u64(_overflow, vdupq_n_u64(0x0), 1);              \
        _a = vorrq_u8(vshrq_n_u64(v, n), _overflow);                        \
}

/* implemented as a macro so we can use vshlq_n_u8 with variable n
void gift_64_vec_sliced_swapmove(uint8x16_t *restrict a, uint8x16_t *restrict b,
                                 const uint8x16_t m, const int n) */
#define gift_64_vec_sliced_swapmove(a, b, m, n)                             \
{                                                                           \
        uint8x16_t _a;                                                      \
        shr(_a, a, n);                                                      \
        const uint8x16_t _t = vandq_u8(veorq_u8(_a, b), m);                 \
        b = veorq_u8(b, _t);                                                \
        shl(_a, _t, n);                                                     \
        a = veorq_u8(a, _a);                                                \
}
\end{lstlisting}

\subsubsection{Round function}
We will examine the round function in closer detail and compare the source code
with the generated assembly.

\begin{lstlisting}[language=c, caption={Round function}]
for (int round = 0; round < ROUNDS_GIFT_64; round++) {
    gift_64_vec_sliced_subcells(s);
    gift_64_vec_sliced_permute(s);

@@

It is obvious to see the compiler has inlined the two function calls to
\texttt{subcells} and \texttt{permute} with lines 1 to 23 originating from the
\texttt{subcells} and lines 24-46 from the \texttt{permute} and round key
addition functions. It has chosen not to unroll the loop, but has moved the
loop counter increment as well as the condition check in between the
\texttt{subcells} instructions to line \ref{lst:cntrincr} and
\ref{lst:cntrchk}. In addition, the loop counter serves a second purpose as an
offset register and is therefore incremented by 0x80=128 instead of just 1.
